Perturbing SU($n$)-symmetric interactions

### 2020-03-10 collective_sun/sun_perturbation.tex
minor changes to text

Our task is essentially to express the multi-body eigenvalue problem
in \eqref{eq:H_VM_psi_eig} in a basis that obeys both permutational
symmetry and translational invariance.  To this end, we first expand
the coupling tensor $\m w$ in a complete basis for symmetric tensors:
\begin{align}
  \m w = \sum_{k\in\C_N\p{M}} w_k \m m_k,
  &&
  \m m_k \equiv \sum_{\sigma\in P_M} \m e_{\sigma\p{k}},
  \label{eq:sym_tensor}
\end{align}
where $P_M$ is the set of all permutations of $M$ symbols, and
$\m e_\ell$ is a unit tensor with one in the entry indexed by $\ell$
and zeros everywhere else.  On a $D$-dimensional lattice of linear
size $L$, translational invariance implies that the coefficients $w_k$
remain the same if all spins indexed by $k$ are shifted by the same
displacement $d\in\ZZ_L^D$.  To account for this invariance, we denote
a translation of all spins in $k$ by a displacement $d\in\ZZ_L^D$ by
$k_d$, and define the translation-symmetrized basis tensors
\begin{align}
  \m m_k^{\t{TS}}
  \equiv \f1{\N_k^{\t{TS}}} \sum_{d\in\ZZ_L^D} \m m_{k_d}
  = \f1{\N_k^{\t{TS}}} \sum_{d\in\ZZ_L^D} \sum_{\sigma\in P_M}
  \m e_{\sigma(k_d)},
\end{align}
where the normalization factor $\N_k^{\t{TS}}$ enforces that the
entries of $\m m_k^{\t{TS}}$ are all either zero or one, which
otherwise might not occur if some translations of $k$ results in the
same choice of spins.  We then define the set $\T_{LD}\p{M}$ of
equivalence classes in $\C_N\p{M}$ under arbitrary translations on
$D$-dimensional lattice of linear size $L$, and expand
\begin{align}
  \m w = \sum_{k\in\T_{LD}\p{M}} w_k \m m_k^{\t{TS}}.
  \label{eq:sym_shift_tensor}
\end{align}
Having expanded $\m w$ in a basis that respects permutational symmetry
and translational invariance, we need to express the eigenvalue
equation \eqref{eq:H_VM_psi_eig} in the same basis.  We therefore
expand
\begin{align}
  \m h\circ\m w
  = \sum_{k\in\T_{LD}\p{M}} \sum_{p\in\ZZ_L^D} \sum_{a=1}^M
  h_{pk_a} w_{k_{ap}} \m m_k^{\t{TS}},
  &&
  \tilde{\m h} * \m w = \sum_{k\in\T_{LD}\p{M}}
  \tilde h_k w_k \m m_k^{\t{TS}},
\end{align}
which implies that the generalized eigenvalue problem in
\eqref{eq:H_VM_psi_eig} can be written as
\begin{align}
  \check{\m h}\c\v{\m w} = \Delta\v{\m w},
\end{align}
where
\begin{align}
  \check{\m h}
  \equiv \sum_{k\in\T_{LD}\p{M}} \sum_{p\in\ZZ_L^D} \sum_{a=1}^M
  h_{pk_a} \op{k}{k_{ap}}
  - \sum_{k\in\T_{LD}\p{M}} \tilde h_k \op{k},
  &&
  \v{\m w} \equiv \sum_{k\in\T_{LD}\p{M}} w_k \ket{k}.
\end{align}
In order to make sense of $\check{\m h}$ as written above, for each
$k\in\T_{LD}\p{M}$ we enforce a canonical choice of spins
$\tilde k\in k\subset\C_N\p{M}$; we define $k_{ap}$ to be the
equivalence class of $\tilde k_{ap}$ (where $\tilde k_{ap}$ equal to
$\tilde k$ with the $a$-th spin $k_a$ replaced by $p$); and we define
$\ket{k}=0$ if $\tilde k$ has any repeated values.  The matrix
$\check{\m h}$ has dimensions
$\abs{\T_{LD}\p{M}}\times\abs{\T_{LD}\p{M}}$, where
$\abs{\T_{LD}\p{M}} \approx { N-1 \choose M-1 }/M \sim N^{M-1}$.

@@

Weak $M$-body perturbations $\V_M$ to SU($n$)-symmetric interactions
can be treated perturbatively, resulting in an effective Hamiltonian
on a permutationally symmetric manifold $\M_0$ of $N$ spins.
Computing this effective Hamiltonian generally requires multiplying
several operators, and projecting the resulting product onto $\M_0$.
Even if the perturbation $\V_M$ is too strong for rigorous
perturbative treatment, as long as it is comparable in magnitude to
the energetic penalty incurred by breaking permutational symmetry, its
effects can still be accounted for by an effective theory on the joint
manifold of permutationally symmetric states together with states of
``weak'' permutational symmetry breaking, e.g.~the manifold of states
generated by the action of $M$-body perturbations on permutationally
symmetric states.  The expansion of operators in a basis for this
joint manifold likewise requires multiplying several operators and
projecting the resulting product onto $\M_0$.  Specifically, if
$\V_M=\sum_{k\in\C_N\p{M}}w_kO_k$, then we can define operators
\begin{align}
  \V_M^{(k)} \equiv \sum_{p\in\C_N\p{M}} w^{(k)}_p O_p,
\end{align}
where $\set{w^{(k)}}$ forms a complete basis for the set of symmetric
tensors of dimension $M$ with zeros on all diagonals (i.e.~zero in all
entries with a repeated index).  Denoting a complete basis of
permutationally symmetric states by $\set{\ket{m}}$, we can then
identify the $M$-body perturbed states
\begin{align}
  \ket{mk} \equiv \f1{\sqrt{\N_{mk}}} \V_M^{(k)} \ket{m},
  &&
  \N_{mk} \equiv \bk{m | \V_M^{(k)} \V_M^{(k)} | m}.
\end{align}
The restriction of a multi-body operator $\Q=\sum_p v_p Q_p$ to this
expanded Hilbert space is then
\begin{align}
  \Q =_M \P_M \Q \P_M,
  &&
  \P_M \equiv \sum_m \op{m} + \sum_{m,k} \op{mk},
\end{align}
where $=_M$ denotes equality up to a restriction to the support of
$\P_M$, and finding matrix elements of $\Q$ now requires computing
products of the form
\begin{align}
  \bk{\ell | \V_M^{(j)} \Q \V_M^{(k)} | m}
  = \sum_{p,q,r} w^{(j)}_p v_q w^{(k)}_r \bk{\ell | O_p Q_q O_r | m}.
  \label{eq:example_prod}
\end{align}
The purpose of this section is to work out such products in a
systematic way.  To this end, we derive a general diagrammatic
expansion of these products, as well as rules to simplify the
resulting expansions so as to minimize their computational cost.
These diagrammatic methods allow us to algorithmically evaluate
products such as \eqref{eq:example_prod} from the bare multi-body
operator data $\p{w^{(j)},O},\p{v,Q},\p{w^{(k)},O}$.  We then work out
some special cases of products that are relevant to the perturbation
theory in Sections \ref{sec:single_body_pert} and
\ref{sec:two_body_pert}.

@@

The process of constructing and simplifying diagrams to compute the
coefficients $w\p{g}$ of the diagrammatic operator product expansion
in \eqref{eq:sym_prod_group} is systematic enough to be converted into
an algorithm for execution on a computer.  The last step necessary to
automate the evaluation of \eqref{eq:sym_prod_group} is expansion of
the multi-spin operators $O\p{g}$ in a basis for the permutationally
symmetric manifold $\M_0$.  Without loss of generality, we consider a
single permutationally symmetric $M$-spin operator $Q$; our task is
essentially to find coefficients of the expansion
\begin{align}
  \P_0 Q \P_0
  = \sum_{a,b} \bk{a|Q|b} \op{a}{b},
\end{align}
where $\set{\ket{a}},\set{\ket{b}}$ are complete bases for $\M_0$, and
the choice of $M$ spins on which $Q$ acts is arbitrary.  In a system
of $N$ spins with $n$ states each, a suitable basis for $\M_0$ can be
labeled by the occupation number of each single-spin state, namely
$a=\p{a_1,a_2,\cdots,a_n}$ with each $a_j\ge 0$ and $\sum_ja_j=N$.
Written out explicitly,
\begin{align}
  \ket{a} = \f1{\sqrt{\C\p{a}}}
  \sum_{\substack{\t{distinct}\\\t{permutations}\\\Pi~\t{of}~\tilde a}}
  \Pi \ket{\tilde a},
  &&
  \C\p{a} \equiv \f{\p{\sum_j a_j}!}{\prod_ja_j!}
  &&
  \ket{\tilde a} \equiv \bigotimes_{j=1}^n \ket{j}^{\otimes a_j},
\end{align}
where the multinomial coefficient $\C\p{a}$ counts the number of
distinct ways to permute the tensor factors of $\ket{\tilde a}$, and
enforces $\bk{a|a}=1$.  We denote the set of all valid assignments of
$S$ spins to $n$ states by $\A_n\p{S}$, such that the standard basis
for the permutationally symmetric manifold of $S$ spins is
$\set{\ket{a}:a\in\A_n\p{S}}$.  Identifying this standard basis allows
us to expand
\begin{align}
  \bk{a|Q|b}
  = \sum_{\substack{\alpha,\beta\in\A_n\p{M}\\\alpha\le a,\,\beta\le b}}
  \delta_{a-\alpha,b-\beta}
  \sqrt{\f{\C\p{\alpha}\C\p{a-\alpha}\C\p{\beta}\C\p{b-\beta}}
    {\C\p{a}\C\p{b}}}
  \bk{\alpha|Q|\beta},
  \label{eq:multi_body_eval}
\end{align}
where the restriction $\alpha\le a$ and difference $a-\alpha$ are
evaluated element-wise, i.e.~$\alpha<a\implies \alpha_j\le a_j$ and
$\p{a-\alpha}_j=a_j-\alpha_j$ for all $j$; and $\delta_{cd}=1$ if
$c=d$ and zero otherwise.  We sum over both $\alpha$ and $\beta$ above
merely to keep the expression symmetric with respect to transposition;
in practice, one can simply sum over $\alpha\in\A_n\p{M}$ and set
$\beta\equiv b-a+\alpha$.  Note that, by slight abuse of notation, the
operator $Q$ on the left of \eqref{eq:multi_body_eval} acts on an
arbitrary choice of $M$ spins (out of $N$), whereas the operator $Q$
on the right of \eqref{eq:multi_body_eval} is simply an $M$-spin
operator, with matrix elements $\bk{\alpha|Q|\beta}$ evaluated with
respect to the $M$-spin states $\ket\alpha,\ket\beta$.

>>>>

Our task is essentially to express the multi-body eigenvalue problem
in \eqref{eq:H_VM_psi_eig} in a basis that obeys both permutational
symmetry and translational invariance.  To this end, we first expand
the coupling tensor $\m w$ in a complete basis for symmetric tensors:
\begin{align}
  \m w = \sum_{k\in\C_N\p{M}} w_k \m m_k,
  &&
  \m m_k \equiv \sum_{\sigma\in P_M} \m e_{\sigma\p{k}},
  \label{eq:sym_tensor}
\end{align}
where $P_M$ is the set of all permutations of $M$ symbols, and
$\m e_\ell$ is a unit tensor with one in the entry indexed by $\ell$
and zeros everywhere else.  On a $D$-dimensional lattice of linear
size $L$, translational invariance implies that the coefficients $w_k$
remain the same if all spins indexed by $k$ are shifted by the same
displacement $d\in\ZZ_L^D$.  To account for this invariance, we denote
a translation of all spins in $k$ by a displacement $d\in\ZZ_L^D$ by
$k_d$, and define the translation-symmetrized basis tensors
\begin{align}
  \m m_k^{\t{TS}}
  \equiv \f1{\N_k^{\t{TS}}} \sum_{d\in\ZZ_L^D} \m m_{k_d}
  = \f1{\N_k^{\t{TS}}} \sum_{d\in\ZZ_L^D} \sum_{\sigma\in P_M}
  \m e_{\sigma(k_d)},
\end{align}
where the normalization factor $\N_k^{\t{TS}}$ enforces that the
entries of $\m m_k^{\t{TS}}$ are all either zero or one, which
otherwise might not occur if some translations of $k$ results in the
same choice of spins.  We then define the set $\T_{LD}\p{M}$ of
equivalence classes in $\C_N\p{M}$ under arbitrary translations on
$D$-dimensional lattice of linear size $L$, and expand
\begin{align}
  \m w = \sum_{k\in\T_{LD}\p{M}} w_k \m m_k^{\t{TS}}.
  \label{eq:sym_shift_tensor}
\end{align}
Having expanded $\m w$ in a basis that respects permutational symmetry
and translational invariance, we need to express the eigenvalue
equation \eqref{eq:H_VM_psi_eig} in the same basis.  We therefore
expand
\begin{align}
  \m h\circ\m w
  = \sum_{k\in\T_{LD}\p{M}} \sum_{p\in\ZZ_L^D} \sum_{a=1}^M
  h_{pk_a} w_{k_{ap}} \m m_k^{\t{TS}},
  &&
  \tilde{\m h} * \m w = \sum_{k\in\T_{LD}\p{M}}
  \tilde h_k w_k \m m_k^{\t{TS}},
\end{align}
which allows us to write the generalized eigenvalue problem in
\eqref{eq:H_VM_psi_eig} as
\begin{align}
  \check{\m h}\c\v{\m w} = \Delta\v{\m w},
\end{align}
where
\begin{align}
  \check{\m h}
  \equiv \sum_{k\in\T_{LD}\p{M}} \sum_{p\in\ZZ_L^D} \sum_{a=1}^M
  h_{pk_a} \op{k}{k_{ap}}
  - \sum_{k\in\T_{LD}\p{M}} \tilde h_k \op{k},
  &&
  \v{\m w} \equiv \sum_{k\in\T_{LD}\p{M}} w_k \ket{k}.
\end{align}
In order to make sense of $\check{\m h}$ as written above, for each
$k\in\T_{LD}\p{M}$ we enforce a canonical choice of spins
$\tilde k\in k\subset\C_N\p{M}$; we define $k_{ap}$ to be the
equivalence class of $\tilde k_{ap}$ (where $\tilde k_{ap}$ is equal
to $\tilde k$ with the $a$-th spin $k_a$ replaced by $p$); and we
define $\ket{k}=0$ if $\tilde k$ has any repeated values.  The matrix
$\check{\m h}$ has dimensions
$\abs{\T_{LD}\p{M}}\times\abs{\T_{LD}\p{M}}$, where
$\abs{\T_{LD}\p{M}} \approx { N-1 \choose M-1 }/M \sim N^{M-1}$.

@@

Weak $M$-body perturbations $\V_M$ to SU($n$)-symmetric interactions
can be treated perturbatively, resulting in an effective Hamiltonian
on a permutationally symmetric manifold $\M_0$ of $N$ spins.
Computing this effective Hamiltonian generally requires multiplying
several operators, and projecting the resulting product onto $\M_0$.
Even if the perturbation $\V_M$ is too strong for rigorous
perturbative treatment, as long as it is comparable in magnitude to
the energetic penalty incurred by breaking permutational symmetry, its
effects can still be accounted for by an effective theory on the joint
manifold of permutationally symmetric states together with states of
``weak'' permutational symmetry breaking, e.g.~the manifold of states
generated by the action of $M$-body perturbations on permutationally
symmetric states.  The expansion of operators in a basis for this
extended manifold likewise requires multiplying several operators and
projecting the resulting product onto $\M_0$.  Specifically, if
$\V_M=\sum_{k\in\C_N\p{M}}w_kO_k$, then we can define operators
\begin{align}
  \V_M^{(k)} \equiv \sum_{p\in\C_N\p{M}} w^{(k)}_p O_p,
\end{align}
where $\set{w^{(k)}}$ forms a complete basis for the set of symmetric
tensors of dimension $M$ with zeros on all diagonals (i.e.~zero in all
entries with a repeated index).  Denoting a complete basis of
permutationally symmetric states by $\set{\ket{m}}$, we can then
identify the $M$-body perturbed states
\begin{align}
  \ket{mk} \equiv \f1{\sqrt{\N_{mk}}} \V_M^{(k)} \ket{m},
  &&
  \N_{mk} \equiv \bk{m | \V_M^{(k)} \V_M^{(k)} | m}.
\end{align}
The restriction of a multi-body operator $\Q=\sum_p v_p Q_p$ to this
expanded Hilbert space is then
\begin{align}
  \Q =_M \P_M \Q \P_M,
  &&
  \P_M \equiv \sum_m \op{m} + \sum_{m,k} \op{mk},
\end{align}
where $=_M$ denotes equality up to a restriction to the support of
$\P_M$, and finding matrix elements of $\Q$ now requires computing
products of the form
\begin{align}
  \bk{\ell | \V_M^{(j)} \Q \V_M^{(k)} | m}
  = \sum_{p,q,r} w^{(j)}_p v_q w^{(k)}_r \bk{\ell | O_p Q_q O_r | m}.
  \label{eq:example_prod}
\end{align}
The purpose of this section is to work out such products in a
systematic way.  To this end, we derive a general diagrammatic
expansion of these products, as well as rules to simplify the
resulting expansions so as to minimize their computational cost.
These diagrammatic methods allow us to algorithmically evaluate
products such as \eqref{eq:example_prod} from the bare multi-body
operator data $\p{w^{(j)},O},\p{v,Q},\p{w^{(k)},O}$.  We then work out
some special cases of products that are relevant to the perturbation
theory in Sections \ref{sec:single_body_pert} and
\ref{sec:two_body_pert}.

@@

The process of constructing and simplifying diagrams to compute the
coefficients $w\p{g}$ of the diagrammatic operator product expansion
in \eqref{eq:sym_prod_group} is systematic enough to be converted into
an algorithm for execution on a computer.  The last step necessary to
automate the evaluation of \eqref{eq:sym_prod_group} is expansion of
the multi-spin operators $O\p{g}$ in a basis for the permutationally
symmetric manifold $\M_0$.  Without loss of generality, we consider a
single permutationally symmetric $M$-spin operator $Q$; our task is
essentially to find coefficients of the expansion
\begin{align}
  \P_0 Q \P_0
  = \sum_{a,b} \bk{a|Q|b} \op{a}{b},
\end{align}
where $\set{\ket{a}},\set{\ket{b}}$ are complete bases for $\M_0$, and
the choice of $M$ spins on which $Q$ acts is arbitrary.  In a system
of $N$ spins with $n$ states each, a suitable basis for $\M_0$ can be
labeled by the occupation number of each single-spin state, namely
$a=\p{a_1,a_2,\cdots,a_n}$ with each $a_j\ge 0$ and $\sum_ja_j=N$.
Written out explicitly,
\begin{align}
  \ket{a} = \f1{\sqrt{\C\p{a}}}
  \sum_{\substack{\t{distinct}\\\t{permutations}\\\Pi~\t{of}~\tilde a}}
  \Pi \ket{\tilde a},
  &&
  \ket{\tilde a} \equiv \bigotimes_{j=1}^n \ket{j}^{\otimes a_j},
  &&
  \C\p{a} \equiv \f{\p{\sum_j a_j}!}{\prod_ja_j!},
\end{align}
where the multinomial coefficient $\C\p{a}$ counts the number of
distinct ways to permute the tensor factors of $\ket{\tilde a}$, and
enforces $\bk{a|a}=1$.  We denote the set of all valid assignments of
$S$ spins to $n$ states by $\A_n\p{S}$, such that the standard basis
for the permutationally symmetric manifold of $S$ spins is
$\set{\ket{a}:a\in\A_n\p{S}}$.  Identifying this standard basis allows
us to expand
\begin{align}
  \bk{a|Q|b}
  = \sum_{\substack{\alpha,\beta\in\A_n\p{M}\\\alpha\le a,\,\beta\le b}}
  \delta_{a-\alpha,b-\beta}
  \sqrt{\f{\C\p{\alpha}\C\p{a-\alpha}\C\p{\beta}\C\p{b-\beta}}
    {\C\p{a}\C\p{b}}}
  \bk{\alpha|Q|\beta},
  \label{eq:multi_body_eval}
\end{align}
where the restriction $\alpha\le a$ and difference $a-\alpha$ are
evaluated element-wise, i.e.~$\alpha<a\implies \alpha_j\le a_j$ and
$\p{a-\alpha}_j=a_j-\alpha_j$ for all $j$; and $\delta_{cd}=1$ if
$c=d$ and zero otherwise.  We sum over both $\alpha$ and $\beta$ above
merely to keep the expression symmetric with respect to transposition;
in practice, one can simply sum over $\alpha\in\A_n\p{M}$ and set
$\beta=b-a+\alpha$.  Note that, by slight abuse of notation, the
operator $Q$ on the left of \eqref{eq:multi_body_eval} acts on an
arbitrary choice of $M$ spins (out of $N$), whereas the operator $Q$
on the right of \eqref{eq:multi_body_eval} is simply an $M$-spin
operator, with matrix elements $\bk{\alpha|Q|\beta}$ evaluated with
respect to the $M$-spin states $\ket\alpha,\ket\beta$.
